Alternative Methods for Implementing Explicit and Finding Implicit Sharing in embedded DSLs

### 2022-12-15 paper/paper.tex
switched from minted to lstlisting

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{listings}
\usepackage{minted}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Christopher Anand \and
Lucas Dutton \and
Steven Gonder
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
TODO The abstract should briefly summarize the contents of the paper in
150--250 words.

@@

\begin{minted}{haskell}
data Exp
  = Add Exp Exp
  | Variable String
  | Constant Int

-- Example
v0 = Variable ``v0''
exp0 = Add v0 (Constant 0)
exp1 = Add exp0 exp0
\end{minted}

@@

\begin{minted}{haskell}
class Exp repr where
  add :: repr Int -> repr Int -> repr Int
  variable :: String -> repr Int
  constant :: Int -> repr Int
\end{minted}

We can then create different instances to implement different functionality.
For example, we can generate the AST from the previous DSL like so
\begin{minted}{haskell}
newtype ExpR a = ExpR { unExpR :: Exp }

instance Exp ExpR where
  constant = ExpR . Constant
  variable = ExpR . Variable
  add (ExpR x) (ExpR y) = ExpR (Add x y)
\end{minted}

Or we can implement pretty printing
\begin{minted}{haskell}
newtype Pretty a = Pretty { unPretty :: String }

instance Exp Pretty where
  add x y = Pretty $ "("++unPretty x++") + ("++unPretty y++")"
  variable x = Pretty x
  constant x = Pretty $ show x
\end{minted}

And use the same DSL code to run either implementation
\begin{minted}{haskell}
exp :: Exp repr => repr Int -> repr Int
exp v0 =
  let
    exp0 = add v0 (constant 0)
  in add exp0 exp0

expR = unExpR $ exp $ variable "v0"
expP = unPretty $ exp $ variable "v0"
\end{minted}

@@

In \cite{kiselyov:sharing}, a solution for detection implicit sharing in finally
tagless style is presented via the method of hash-consing. You can find a more
throrough explanation of the method there, but we'll give an overview here. This
method first involves defining a DAG type, for example
\begin{minted}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int

@@

newtype DAG = DAG (BiMap Node) deriving Show
\end{minted}

@@

\begin{minted}{haskell}
newtype Graph a = Graph { unGraph :: State DAG NodeID }

instance Exp Graph where
  constant x = Graph (hashcons $ NConstant x)
  variable x = Graph (hashcons $ NVariable x)
  add e1 e2 = Graph (do
                     h1 <- unGraph e1
                     h2 <- unGraph e2
                     hashcons $ NAdd h1 h2)
\end{minted}

The trick to uncovering sharing in the implementation is implemented via the
{\bf hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{minted}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{minted}

@@

When we wrap our State monad in finally tagless style, we lose some of Haskell's
built-in sharing capability. Consider the following code, note that the use of
local variables explicitely defines the computation $x + y$ to only be computed
once
\begin{minted}{haskell}
haskellSharing x y=
 let
   z = x + y
 in z + z
\end{minted}

Implicit sharing via hash-consing prevents duplication in the resulting DAG, but
unfortunately doesn't prevent redundant computation. Consider the following
equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
\begin{minted}{haskell}
dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
dslSharing x y =
  let
    z = add x y
  in add z z
\end{minted}
Note {\bf z} is a wrapper around a State monad. Recall the implementation of
{\bf add} via hash consing
\begin{minted}{haskell}
add e1 e2 = Graph (do
                   h1 <- unGraph e1
                   h2 <- unGraph e2
                   hashcons $ NAdd h1 h2)
\end{minted}
The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
State monad, meaning even if {\bf e1} and {\bf e2} are the same shared Haskell
value, their underlying computations will be performed twice. Hash-consing will
prevent these redundencies from appearing in the resulting DAG, but the entire
unshared AST wil still be traversed, performing a hash-cons on each node.

Consider a chain of {\bf add}'s  with sharing, for example
\begin{minted}{haskell}
addChains :: Exp repr => Expr Int -> Expr Int
addChains x0 = let
    x1 = add x0 x0
    x2 = add x1 x1
    ...
  in xn
\end{minted}

@@

\cite{kiselyov:sharing} acknowledges the amount of computation with hash-consing
can get out of control, and proposes an ad-hoc solution, explicit sharing via a
custom let construct
\begin{minted}{haskell}
class ExpLet repr where
  let_ :: repr a -> (repr a -> repr b) -> repr b
instance ExpLet Graph where
  let_ e f = Graph (do x <- unGraph e
                     unGraph $ f (Graph (return x)))

addChains x =
  let_ x (\x0 ->
  let_ (add x0 x0)  (\x1 ->
  let_ (add x1 x1)  (\x2 ->
   ...
  )))
\end{minted}
This makes the code a bit clunky and adds an extra burden on the DSL writer, but
it prevents unnecessary hash-consing in our example. However the method has it's
limitations, suppose we want to write a DSL function that returns multiple
outputs, such as tuples or container types like lists (for example
\mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr Int)}). %TODO
                                % need to give implementation of this function?
First of all, we need to implement different versions of the custom let construct to
correspond to the number of outputs
\begin{minted}{haskell}
class ExpLet repr where
  let_ :: repr a -> (repr a -> repr b) -> repr b
  let_ 2 :: repr a -> (repr a -> (repr b,repr c)) -> (repr b,repr c)

instance ExpLet repr where
   ...
   let_2 e f = let
       g0 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o0)
       g1 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o1)
       in (g0,g1)
\end{minted}
If need be, its possible to enumerate custom let instances for every amount of
outputs or container types we would need, we could even use template Haskell to
accomplish this. However, this custom let construct now has a new source of
redundancy in its outputs. Each output it returns will now have to individually
evaluate it's input, so a chain of DSL functions that output 2 or more values
will suffer from the same exponential scaling of hashcons.

@@

\begin{minted}{haskell}
data Graph a = Graph { unGraph :: State DAG NodeID
                     , stringAST :: ByteString }

data DAG = DAG { Trie (Node,NodeID)
               } deriving Show
\end{minted}

The Trie essentially serves as our new BiMap, the resulting DAG contained in its
lookup values. We essentially still perform the hash-consing technique but using
the AST to perform the lookup
\begin{minted}{haskell}
hashcons :: ByteString -> Node -> State DAG NodeID
hashcons sAST node = do
 DAG trie maxID <- get
 case Trie.lookup sAST trie of
   Nothing -> let maxID' = maxID+1
                  trie' = Trie.insert sAST (node,maxID+1) trie
               in do put $ DAG trie' maxID'
                     return maxID'
   Just (_,nodeID) -> return nodeID

@@

\end{minted}

The instance implementations for \mintinline{haskell}{constant} and
\mintinline{haskell}{variable} work roughly the same, the novelty of the method
is in how we handle DSL functions that take other DSL State as input like
\mintinline{haskell}{add}. First we need to construct a byte string AST from
it's input ASTs, there's a lot of ways we could go about this to attempt to
minimize memory. A naive implementation would look similar to a pretty printer.
Then we need to sequence it's inputs without evaluating the inner state if
unnecessary. We do this through the implementation of \mintinline{haskell}{seqArgs}
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to look up its corresponding byte string AST in the Trie. This will prevent
redundant hashconsing without the need for explicit sharing. However this method
suffers from its own drawbacks.

@@

\section{Explicit Sharing Of ByteString ASTs}
We propose another solution to this issue, taking inspiration again from the
\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
sharing. This time, the construct will substitute the current byte string for a
more compact label. For safety purposes, we need to keep track of a table of
these labels and their corresponding ASTs, to make sure we don't insert of the
same labels.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagCacheMap :: Map ByteString ByteString
               } deriving Show

@@

-- TODO better example?
test x y = let
  z = cache "z" (add x y)
  in add z z
\end{minted}

The cache operation replaces the current byte string AST with a new label, and
we'll define a new operation \mintinline{haskell}{runCache} that will check if
the label already exists in the cache map before inserting it.
\begin{minted}{haskell}
runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
runCache sAST mAddCache cacheMap = do
  case mAddCache of
    Nothing -> return ()
    Just sAST0 ->
      case Trie.lookup sAST cacheMap of
        Nothing -> let cacheMap' = Trie.insert sAST sAST0 cacheMap
                   in modify (\dag -> dag { dagCache = cacheMap' })
        Just sAST1 -> if sAST1 == sAST0
                         then return ()
                         else error $ "attempted to recache: " ++ show sAST

seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST mCache) =
      do DAG trie cacheMap _ <- get
         runCache sAST mCache cacheMap
         ...
\end{minted}
We need to make sure we don't attempt to insert the same cache label for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the State monad to prevent or modify the subtitution. The best we can do is
crash the program, or if we use a monad tansformer, we ould use
Control.Monad.Except to through an exception. Either way it's up to the DSL
writer to insure they don't reuse the same label.

@@

% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
% Local Variables:
% LaTeX-verbatim-environments-local: ("minted")
% eval: (setq-local LaTeX-indent-environment-list (cons '("minted" current-indentation) (default-value 'LaTeX-indent-environment-list)))
% End:

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{listings}
% \usepackage{lstlisting}

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Christopher Anand \and
Lucas Dutton \and
Steven Gonder
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
TODO The abstract should briefly summarize the contents of the paper in
150--250 words.

@@

\begin{lstlisting}{haskell}
data Exp
  = Add Exp Exp
  | Variable String
  | Constant Int

-- Example
v0 = Variable ``v0''
exp0 = Add v0 (Constant 0)
exp1 = Add exp0 exp0
\end{lstlisting}

@@

\begin{lstlisting}{haskell}
class Exp repr where
  add :: repr Int -> repr Int -> repr Int
  variable :: String -> repr Int
  constant :: Int -> repr Int
\end{lstlisting}

We can then create different instances to implement different functionality.
For example, we can generate the AST from the previous DSL like so
\begin{lstlisting}{haskell}
newtype ExpR a = ExpR { unExpR :: Exp }

instance Exp ExpR where
  constant = ExpR . Constant
  variable = ExpR . Variable
  add (ExpR x) (ExpR y) = ExpR (Add x y)
\end{lstlisting}

Or we can implement pretty printing
\begin{lstlisting}{haskell}
newtype Pretty a = Pretty { unPretty :: String }

instance Exp Pretty where
  add x y = Pretty $ "("++unPretty x++") + ("++unPretty y++")"
  variable x = Pretty x
  constant x = Pretty $ show x
\end{lstlisting}

And use the same DSL code to run either implementation
\begin{lstlisting}{haskell}
exp :: Exp repr => repr Int -> repr Int
exp v0 =
  let
    exp0 = add v0 (constant 0)
  in add exp0 exp0

expR = unExpR $ exp $ variable "v0"
expP = unPretty $ exp $ variable "v0"
\end{lstlisting}

@@

In \cite{kiselyov:sharing}, a solution for detection implicit sharing in finally
tagless style is presented via the method of hash-consing. You can find a more
throrough explanation of the method there, but we'll give an overview here. This
method first involves defining a DAG type, for example
\begin{lstlisting}{haskell}
type NodeID = Int
data Node = NAdd NodeID NodeID
          | NVariable String
          | NConstant Int

@@

newtype DAG = DAG (BiMap Node) deriving Show
\end{lstlisting}

@@

\begin{lstlisting}{haskell}
newtype Graph a = Graph { unGraph :: State DAG NodeID }

instance Exp Graph where
  constant x = Graph (hashcons $ NConstant x)
  variable x = Graph (hashcons $ NVariable x)
  add e1 e2 = Graph (do
                     h1 <- unGraph e1
                     h2 <- unGraph e2
                     hashcons $ NAdd h1 h2)
\end{lstlisting}

The trick to uncovering sharing in the implementation is implemented via the
{\bf hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{lstlisting}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{lstlisting}

@@

When we wrap our State monad in finally tagless style, we lose some of Haskell's
built-in sharing capability. Consider the following code, note that the use of
local variables explicitely defines the computation $x + y$ to only be computed
once
\begin{lstlisting}{haskell}
haskellSharing x y=
 let
   z = x + y
 in z + z
\end{lstlisting}

Implicit sharing via hash-consing prevents duplication in the resulting DAG, but
unfortunately doesn't prevent redundant computation. Consider the following
equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
\begin{lstlisting}{haskell}
dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
dslSharing x y =
  let
    z = add x y
  in add z z
\end{lstlisting}
Note {\bf z} is a wrapper around a State monad. Recall the implementation of
{\bf add} via hash consing
\begin{lstlisting}{haskell}
add e1 e2 = Graph (do
                   h1 <- unGraph e1
                   h2 <- unGraph e2
                   hashcons $ NAdd h1 h2)
\end{lstlisting}
The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
State monad, meaning even if {\bf e1} and {\bf e2} are the same shared Haskell
value, their underlying computations will be performed twice. Hash-consing will
prevent these redundencies from appearing in the resulting DAG, but the entire
unshared AST wil still be traversed, performing a hash-cons on each node.

Consider a chain of {\bf add}'s  with sharing, for example
\begin{lstlisting}{haskell}
addChains :: Exp repr => Expr Int -> Expr Int
addChains x0 = let
    x1 = add x0 x0
    x2 = add x1 x1
    ...
  in xn
\end{lstlisting}

@@

\cite{kiselyov:sharing} acknowledges the amount of computation with hash-consing
can get out of control, and proposes an ad-hoc solution, explicit sharing via a
custom let construct
\begin{lstlisting}{haskell}
class ExpLet repr where
  let_ :: repr a -> (repr a -> repr b) -> repr b
instance ExpLet Graph where
  let_ e f = Graph (do x <- unGraph e
                     unGraph $ f (Graph (return x)))

addChains x =
  let_ x (\x0 ->
  let_ (add x0 x0)  (\x1 ->
  let_ (add x1 x1)  (\x2 ->
   ...
  )))
\end{lstlisting}
This makes the code a bit clunky and adds an extra burden on the DSL writer, but
it prevents unnecessary hash-consing in our example. However the method has it's
limitations, suppose we want to write a DSL function that returns multiple
outputs, such as tuples or container types like lists (for example
\lstinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr Int)}). %TODO
                                % need to give implementation of this function?
First of all, we need to implement different versions of the custom let construct to
correspond to the number of outputs
\begin{lstlisting}{haskell}
class ExpLet repr where
  let_ :: repr a -> (repr a -> repr b) -> repr b
  let_ 2 :: repr a -> (repr a -> (repr b,repr c)) -> (repr b,repr c)

instance ExpLet repr where
   ...
   let_2 e f = let
       g0 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o0)
       g1 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o1)
       in (g0,g1)
\end{lstlisting}
If need be, its possible to enumerate custom let instances for every amount of
outputs or container types we would need, we could even use template Haskell to
accomplish this. However, this custom let construct now has a new source of
redundancy in its outputs. Each output it returns will now have to individually
evaluate it's input, so a chain of DSL functions that output 2 or more values
will suffer from the same exponential scaling of hashcons.

@@

\begin{lstlisting}{haskell}
data Graph a = Graph { unGraph :: State DAG NodeID
                     , stringAST :: ByteString }

data DAG = DAG { Trie (Node,NodeID)
               } deriving Show
\end{lstlisting}

The Trie essentially serves as our new BiMap, the resulting DAG contained in its
lookup values. We essentially still perform the hash-consing technique but using
the AST to perform the lookup
\begin{lstlisting}{haskell}
hashcons :: ByteString -> Node -> State DAG NodeID
hashcons sAST node = do
 DAG trie maxID <- get
 case Trie.lookup sAST trie of
   Nothing -> let maxID' = maxID+1
                  trie' = Trie.insert sAST (node,maxID+1) trie
               in do put $ DAG trie' maxID'
                     return maxID'
   Just (_,nodeID) -> return nodeID

@@

\end{lstlisting}

The instance implementations for \lstinline{haskell}{constant} and
\lstinline{haskell}{variable} work roughly the same, the novelty of the method
is in how we handle DSL functions that take other DSL State as input like
\lstinline{haskell}{add}. First we need to construct a byte string AST from
it's input ASTs, there's a lot of ways we could go about this to attempt to
minimize memory. A naive implementation would look similar to a pretty printer.
Then we need to sequence it's inputs without evaluating the inner state if
unnecessary. We do this through the implementation of \lstinline{haskell}{seqArgs}
\begin{lstlisting}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{lstlisting}

We only evaluate the inner state \lstinline{haskell}{sT} of each argument if we
fail to look up its corresponding byte string AST in the Trie. This will prevent
redundant hashconsing without the need for explicit sharing. However this method
suffers from its own drawbacks.

@@

\section{Explicit Sharing Of ByteString ASTs}
We propose another solution to this issue, taking inspiration again from the
\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
sharing. This time, the construct will substitute the current byte string for a
more compact label. For safety purposes, we need to keep track of a table of
these labels and their corresponding ASTs, to make sure we don't insert of the
same labels.
\begin{lstlisting}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagCacheMap :: Map ByteString ByteString
               } deriving Show

@@

-- TODO better example?
test x y = let
  z = cache "z" (add x y)
  in add z z
\end{lstlisting}

The cache operation replaces the current byte string AST with a new label, and
we'll define a new operation \lstinline{haskell}{runCache} that will check if
the label already exists in the cache map before inserting it.
\begin{lstlisting}{haskell}
runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
runCache sAST mAddCache cacheMap = do
  case mAddCache of
    Nothing -> return ()
    Just sAST0 ->
      case Trie.lookup sAST cacheMap of
        Nothing -> let cacheMap' = Trie.insert sAST sAST0 cacheMap
                   in modify (\dag -> dag { dagCache = cacheMap' })
        Just sAST1 -> if sAST1 == sAST0
                         then return ()
                         else error $ "attempted to recache: " ++ show sAST

seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST mCache) =
      do DAG trie cacheMap _ <- get
         runCache sAST mCache cacheMap
         ...
\end{lstlisting}
We need to make sure we don't attempt to insert the same cache label for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the State monad to prevent or modify the subtitution. The best we can do is
crash the program, or if we use a monad tansformer, we ould use
Control.Monad.Except to through an exception. Either way it's up to the DSL
writer to insure they don't reuse the same label.

@@

% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
% Local Variables:
% LaTeX-verbatim-environments-local: ("lstlisting")
% eval: (setq-local LaTeX-indent-environment-list (cons '("lstlisting" current-indentation) (default-value 'LaTeX-indent-environment-list)))
% End:
